Discrete Quantum Theories and Computing

### 2018-12-14 dissertation.tex
dissertation.lyx "Chapter 1 Introduction" Updating...

Since we as human cannot distinguish quantities differing only by
an arbitrarily small extent \cite{Turing_1937,Gisin2017}, the quantum
mechanical model based on the continuum of complex numbers may not
capture our ability of predicting quantum phenomenon and the power
of quantum computation. Being agnostic about whether the reality is
ultimately discrete or continuous, we considered two types of quantum
models to incorporate our limitation of distinguishability. By assuming
the quantum states are ultimately quantized and distinguishable, our
first type of quantum theories replaces the complex numbers by discrete
finite fields. We then focus on the limitation of distinguishability
itself, incorporate the finite precision measurement into our model,
and consider quantum interval-valued probability measures.

After first suggesting by Richard Feynman 40 years ago~\cite{Feynman1982Simulating},
IBM, \textitalian{Rigetti}, Google, Microsoft, and many other companies
and countries start to build realistic quantum computers recently
\cite{Shieber2018,Kahn2019}, and seek more efficient ways to simulate
chemical molecules and to break down RSA \cite{boneh1999twenty,wiki:RSAFactoring}.
No matter whether we use classical or quantum computers to solve these
two tasks, we notice that they require two different types of quantities.
The quantities required by RSA are discrete, and it is beyond our
ability to distinguish whether the quantities required by simulating
molecules are discrete or not. Before we dive into the discussion
of these quantities in quantum computation, it is instructive to review
how classical quantities are characterized and computed.

Discrete quantities can be easily characterized by whole numbers~\cite{ParkerBaldridge2004}.
Even if the proportion of whole numbers are fractions, the operations
among fractions are essentially the same as those among whole numbers
over the common denominator. In general, discrete computation can
be reliably repeated with exact equivalence, and any digital computer
can faithfully simulate a Turing machine limited only by the available
memory and time~\cite{Turing_1937}.

In contrast, the non-apparently discrete quantities confused many
cultures since very beginning. On one hand, the ancient Chinese philosopher
Chuang-tzu (庄子) didn't believe these quantities have a minimum unit,
and expressed ``If a one-foot-long stick is cut into halves every
day, the cutting will never come to an end''\footnote{``一尺之棰，日取其半，萬世不竭。'' in Chinese \cite{Zhuangzi1999,Liu2018}.}.
On the other hand, Greek mathematicians in the school of Pythagoras
believed every mathematical model must be ultimately discrete. They
believed so much that they in legend murdered Hippasus because he
found there is no minimum unit between a side of a square and its
diagonal. Comparing the idea that the length of a square's diagonal
is not a physical quantity or there is no square physically, working
with a mathematical model without minimum unit seems more appealing.
Since the mathematical model is continuous, their corresponding physical
quantities are also assumed to be continuous. Later, these continuous
quantities becomes the basic component of analog computers. After
logarithm was discovered, a slide rule, also known as a slipstick,
is used to compute multiplication, division, and more complex operations
as a mechanical analog computer~\cite{wiki:SlideRule}. Charge amplifiers
is used to build electrical analog computers and compute the integration
in calculus \cite{wiki:Integrator,wiki:ChargeAmplifier}.

@@

Contextuality & Yes & Supercontextuality &  &  &  & %
Yes when measuring\tabularnewline[\doublerulesep]
precise enough\tabularnewline[\doublerulesep]

@@

\todo{Introduction Cleaned up until here.}

\todo{Write a story so that physicists will also be interested in
this thesis. For example, quantum simulation problem and how to apply
these to other physicists work (say two more pages?). Check our previous
papers. Explicitly say what question we want to solve. In particular,
merging quantum mechanics and computer science is too broad for a
thesis. }

\todo{Maybe start with this kind of sentence “We want a more faithful
quantum theory not only because the current quantum theory don't take
into account of the measurement precision, but also because its computational
model doesn't fit the current development of computer science.”
}\todo{See if we can cite this paper~\cite{Gisin2017} for finite
precision measurement. }

\todo{新的大綱：主旨：probability amplitude到底是離散還是連續的？
\begin{itemize}
\item Some quantities are obvious discrete, but how to model other quantities
which we don't know?
\begin{itemize}
\item 一尺之槌，日取其半，萬世不竭。
\item Zeno 追龜的故事
\item $\sqrt{2}$是無理數謀殺案(物理量？)
\item Euclid?(物理量？)
\end{itemize}
當我們的儀器不足以分辨出最小單位時，假設沒有最小單位很合理，而且可以很好的解釋我們觀察到的現象。
\item 很多大家本來相信是連續的物理量最後被證明是離散的：
\begin{itemize}
\item “一尺之槌，日取其半，萬世不竭。”是錯的。
\item 光電效應
\item 角動量(角度？)
\end{itemize}
\end{itemize}
}

The marriage of quantum mechanics and computer science first envisioned
and popularized by Feynman has created an awkward, but opportune,
moment. This embarrassing dilemma was concisely summarized by Prof.~Aaronson
that one of the the following three statements is false \cite{AaronsonArkhipov2011,BFRDARW2013}:
\begin{enumerate}[label=(\roman{enumi})]
\item \label{enu:Textbook-quantum-mechanics}Textbook quantum mechanics
is correct. 
\item \label{enu:no-efficient-classical-factoring}There does not exist
an efficient classical factoring algorithm. 
\item \label{enu:extended-Church-Turing-Thesis}The extended Church-Turing
thesis — that probabilistic Turing machines can efficiently simulate
any physically realizable model of computation — is correct \cite{Bernstein:1993:QCT:167088.167097,Kaye2007}. 
\end{enumerate}
There is overwhelming evidence to support each of these statements.
The theoretical framework of quantum mechanics \ref{enu:Textbook-quantum-mechanics}
has withstood decades of experimental confirmation. Entire industries
are founded on the assumption \ref{enu:no-efficient-classical-factoring}
that algorithms like RSA are secure and they also have withstood years
of attempted attacks \cite{boneh1999twenty,wiki:RSAFactoring}. Finally
the entire field of complexity theory in computer science which has
also withstood years of field testing rests, in essence, on assumption
\ref{enu:extended-Church-Turing-Thesis} \cite{Aaronson2005,Piccinini2015}.
And yet at least one of these three statements \emph{must be false}!
Indeed if there is a corresponding efficient classical factoring algorithm,
then we concede \ref{enu:no-efficient-classical-factoring}. If it
is correct that Shor's efficient factoring algorithm is realizable,
and we can prove there is no efficient classical factoring algorithm,
then we concede \ref{enu:extended-Church-Turing-Thesis}. Otherwise,
if we cannot implement Shor's algorithm no matter how hard we try,
textbook quantum mechanics, i.e., \ref{enu:Textbook-quantum-mechanics}
may need to be improved by a better theory. It is unlikely that there
will be a simple resolution to this awkward situation. It is more
likely that the resolution will emerge from deep and careful analyses
of the foundations of each field.

When we check the compatibility between quantum mechanics and computer
science, we found one of the biggest difference is that the success
of digital computers is based on discrete computation while quantum
physics is usually based on the continuum of real and complex numbers.
This suggests whether the computational model is discrete or continuous
may be another important factor parallel to \todo{Find a phrase better
than ``parallel to''. }whether the model is classical or quantum.
\todo{Add discussion about randomness, like: ``Another big difference
is that randomness in quantum computing is inherited from its underlying
quantum physics while randomness in probabilistic Turing machine is
an addition feature acquired by adding random bits to a Turing machine.''
}Although building a realistic quantum computer is an ongoing project~\cite{Shieber2018},
reviewing the challenge posed by classical discrete and continuous
models might provide what we are facing when moving on to the quantum
world.

There was a long history that human believe they can compute with
the continuum of real numbers. If the length of a line segment represents
a real number, Euclid thought he could use an unmarked ruler and a
compass to compute the real numbers about 300 B.C~\cite{MCC2006}.
After logarithm was discovered, a slide rule, also known as a slipstick,
is used to compute multiplication, division, and more complex operations
as a mechanical analog computer~\cite{wiki:SlideRule}. Although
more and more complex analog computers, like integrators \cite{wiki:Integrator,wiki:ChargeAmplifier},
is used, there is still a gap between the computation preformed in
an analog computer and the idealized operations among real numbers.
One problem is that the quality computed by an analog computer might
actually be quantized. For example, charge amplifier is believed to
produce the integrated value of the input current~\cite{wiki:ChargeAmplifier}
although the input charge must actually be an integer multiple of
the elementary charge~\cite{wiki:ElementaryCharge}. Even for the
qualities which doesn't known to be quantized, the observed computational
power of an analog computer might still be weaker than the theoretical
prediction because the precision of an analog computer is limited
generally three or four significant figures~\cite{wiki:AnalogComputer}.
For example, Blum, Cucker, Shub, and Smale modeled their machine to
allow the operations of deciding whether a number is greater or equal
to zero over real numbers and whether a number is exactly zero over
complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}. Since
no physical analog computer could branch in these operations based
on a few significant figures, the theoretical prediction of Blum-Cucker-Shub-Smale
(BCSS) machine cannot fully apply on physical devices.

Being aware that symbols within a computation should not differ to
an arbitrarily small extent~\cite{Turing_1937}, Turing defined his
machine with only finite number of symbols and configurations. One
of the benefit of discrete computation is that it can be reliably
repeated with exact equivalence. Therefore, any digital computer can
faithfully simulate a Turing machine limited only by the available
memory and time.

\begin{table}
\noindent \centering{}\caption{\label{tab:Comparison}Comparison between the classical computational
models and their physical realizations\todo{Put the third row in
the middle alignment.}}
\begin{tabularx}{\linewidth}{ccX}
\toprule 
\addlinespace
 & Discrete & Continuum\tabularnewline\addlinespace
\midrule
\addlinespace
\midrule
\addlinespace
Theoretical  Model & Turing machine & BCSS machine\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
Physical Realization & Digital Computer & Analog Computer\tabularnewline\addlinespace
\addlinespace
\midrule
\addlinespace
How the models realize? & Reliably & Not Reliably:
\begin{itemize}
\item The quality might be quantized.
\item The precision of an analog computer is low.
\end{itemize}\tabularnewline\addlinespace
\bottomrule
\addlinespace
\end{tabularx}
\end{table}
The situations for classical computation is summarized in Table~\ref{tab:Comparison},
and similar situations could potentially be applied to quantum computation.
\todo{Introduction cleaning up until here. }Similar to classical
analog computers manipulating, say, the position of substance, the
most widely used model of quantum computing, the quantum circuit model,
manipulates the probability amplitudes to do the computation. \todo{Play
with some cloud quantum computers to see if I can find more support
evidence. }Because we are agnostic about whether probability amplitude
is quantized or can really be valued in the continuum of complex numbers,
we tried to develop two different types of models: the quantum theories
and computing over finite fields are discrete models, and the quantum
interval-valued probability measures incorporate the idea of measurement
precision into the quantum theory. To distinguish the modified quantum
theories from the usual ones, the usual ones will be called the ``conventional
quantum theory'' where necessary and it will be reviewed in Chapter~\ref{chap:Conventional-Quantum-Theory}~\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.}.

\todo{Beginning of Merge Left}Since conventional probability amplitude
is in the field of complex numbers, and the field structure is essential
to compute the probability from some probability amplitudes, we in
Chapter~\ref{chap:DQT=000026DQC} tried to build discrete quantum
theories and computing by considering them over finite fields.

>>>>

Since we as human cannot distinguish quantities differing only by
an arbitrarily small extent \cite{Turing_1937,Gisin2017}, the quantum
mechanical model based on the continuum of complex numbers may not
capture our ability of predicting quantum phenomenon and the power
of realistic quantum computers. To incorporate our limitation of distinguishability,
but agnostic about whether the reality is ultimately discrete or continuous,
we considered two types of quantum models. By assuming the quantum
states are ultimately discrete and distinguishable, our first type
of quantum theories replaces the complex numbers by finite fields.
We then focus on the limitation of distinguishability itself, incorporate
the finite precision measurement into our model, and consider quantum
interval-valued probability measures.

After first suggested by Richard Feynman 40 years ago~\cite{Feynman1982Simulating},
IBM, \textitalian{Rigetti}, Google, Microsoft, and many other companies
and countries start to build realistic quantum computers recently
\cite{Shieber2018,Kahn2019}, and seek more efficient ways to simulate
chemical molecules and to break down RSA \cite{boneh1999twenty,wiki:RSAFactoring}.
No matter whether we use classical or quantum computers to solve these
two tasks, they require two different types of quantities. The question
and answer of RSA are both discrete integers, but it is beyond our
ability to distinguish whether the quantities required by simulating
molecules are discrete or not. Before we dive into the discussion
of these quantities in quantum computation, it is instructive to review
how classical quantities are characterized and computed.

Discrete quantities can be easily characterized by integers~\cite{ParkerBaldridge2004}.
Even if the proportion of integers are fractions, the operations among
fractions are essentially the same as those among integers over the
common denominator. In general, discrete computation can be reliably
repeated with exact equivalence, and any digital computer can faithfully
simulate a Turing machine limited only by the available memory and
time~\cite{Turing_1937}.

In contrast, the non-apparently discrete quantities confused many
cultures since very beginning. For example, the ancient Chinese philosopher
Chuang-tzu (庄子) didn't believe these quantities have a minimum unit,
and expressed ``If a one-foot-long stick is cut into halves every
day, the cutting will never come to an end''\footnote{``一尺之棰，日取其半，萬世不竭。'' in Chinese \cite{Zhuangzi1999,Liu2018}.}.
As another example, Greek mathematicians in the school of Pythagoras
believed every mathematical model must be ultimately discrete, and
they in legend murdered Hippasus because he proved there is no minimum
unit between a side of a square and its diagonal. Comparing the idea
that the length of a square's diagonal is not a physical quantity
or there is no square physically, working with a mathematical model
without minimum unit seems more appealing. Since the mathematical
model is continuous, their corresponding physical quantities are naturally
assumed to be continuous. Later, these continuous quantities becomes
the basic component of analog computers. After logarithm was discovered,
a slide rule, also known as a slipstick, is used to compute multiplication,
division, and more complex operations as a mechanical analog computer~\cite{wiki:SlideRule}.
Charge amplifiers are used to build electrical analog computers and
compute the integration in calculus \cite{wiki:Integrator,wiki:ChargeAmplifier}.

Although assuming physical quantities is as continuous as their mathematical
models are convenience, the physical quantities discussed in previous
paragraph are not simply real numbers. Chuang-tzu's stick are made
by molecules and cannot be cut in halves endlessly. The input charge
of an integrator an must actually be an integer multiple of the elementary
charge. A prefect square cannot be draw physically. It is beyond our
ability to decide whether a prefect square exists in the space because
whether space itself is ultimately discrete or continuous is still
an solved question. Even if it exists, human have no ability to distinguish
them from an almost prefect square differing only by an arbitrarily
small extent. Similar situation applying to the length of a slide
rule, it is beyond our ability to decide whether its length is ultimately
discrete or continuous. Even if it is continuous, there is no way
to identify which real number represents the length of a slide rule.
As the results, even if analog computers really store real numbers,
they cannot be read, write, and use to its exact value to branch the
computation. The opposite of the last condition sometimes is assumed
by real computational models. For example, Blum, Cucker, Shub, and
Smale modeled their BCSS machine to allow the operations of deciding
whether a number is greater or equal to zero over real numbers and
whether a number is exactly zero over complex numbers \cite{BSS1989,Ziegler2007,blum2012complexity}.
Therefore, the computational power of analog computers is far weaker
than their theoretical prediction.

\begin{table}
\caption{\label{tab:Comparison}Comparison between the classical quantities
and their mathematical models}

\centering{}%
\begin{tabular}{cccc}
\toprule 
\addlinespace
Physical quantities & Discrete & %
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Discrete with\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
extreme small unit\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & We don't know\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational devices & Digital computer & Analog computer & Analog computer\tabularnewline\addlinespace
\midrule
\addlinespace
\addlinespace
\begin{tabular}{c}
\noalign{\vskip\doublerulesep}
Mathematical\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\noalign{\vskip\doublerulesep}
representation\tabularnewline[\doublerulesep]
\noalign{\vskip\doublerulesep}
\end{tabular} & Integers & Real numbers & Real numbers\tabularnewline\addlinespace
\addlinespace
\addlinespace
Computational model  & Turing machine & BCSS machine & BCSS machine\tabularnewline\addlinespace
\midrule
\addlinespace
\addlinespace
How the models realize? & Reliably & Not reliably & Not reliably\tabularnewline\addlinespace
\bottomrule
\addlinespace
\end{tabular}
\end{table}
The situations for classical quantities and the computational power
above them are summarized in Table~\ref{tab:Comparison}, and similar
situations could potentially be applied to quantum mechanics and computation.
The physical quantity manipulated by the quantum circuit model, the
most widely used model of quantum computing, is the probability amplitudes
of states. \todo{Introduction Cleaned up until here. }Because we
are agnostic about whether probability amplitude is quantized or can
really be valued in the continuum of complex numbers, we tried to
develop two different types of models: the quantum theories and computing
over finite fields are discrete models, and the quantum interval-valued
probability measures incorporate the idea of measurement precision
into the quantum theory. To distinguish the modified quantum theories
from the usual ones, the usual ones will be called the ``conventional
quantum theory'' where necessary and it will be reviewed in Chapter~\ref{chap:Conventional-Quantum-Theory}~\footnote{Alternative terminology in the literature includes ``actual,'' ``standard,''
and ``ordinary'' quantum theory.}.

@@

@@

\todo{Beginning of Merge Left}Since conventional probability amplitude
is in the field of complex numbers, and the field structure is essential
to compute the probability from some probability amplitudes, we in
Chapter~\ref{chap:DQT=000026DQC} tried to build discrete quantum
theories and computing by considering them over finite fields.
